---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# easyclimate: Easy access to high-resolution daily climate data for Europe

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/VeruGHub/easyclimate/workflows/R-CMD-check/badge.svg)](https://github.com/VeruGHub/easyclimate/actions)
[![Codecov test coverage](https://codecov.io/gh/VeruGHub/easyclimate/branch/master/graph/badge.svg)](https://codecov.io/gh/VeruGHub/easyclimate?branch=master)
<!-- badges: end -->

Easily get high-resolution (1 km) daily climate data (precipitation, minimum and maximum temperatures) across Europe, from the European climatic database ftp://palantir.boku.ac.at/Public/ClimateData/.


## Installation

``` r
# install.packages("remotes")
remotes::install_github("VeruGHub/easyclimate")
```


## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
# library(easyclimate)
## basic example code
```


## CITATION

```{r echo=FALSE}
citation("easyclimate")
```



