---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)

```

# `easyclimate`

# Easy access to high-resolution daily climate data for Europe

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/VeruGHub/easyclimate/workflows/R-CMD-check/badge.svg)](https://github.com/VeruGHub/easyclimate/actions)
[![Codecov test coverage](https://codecov.io/gh/VeruGHub/easyclimate/branch/master/graph/badge.svg)](https://codecov.io/gh/VeruGHub/easyclimate?branch=master)
<!-- badges: end -->


Get high-resolution (1 km) daily climate data (precipitation, minimum and maximum temperatures) for Europe from the European climatic database hosted at ftp://palantir.boku.ac.at/Public/ClimateData/. Data are available from 1950 to 2017.

This climatic dataset was originally built by [A. Moreno & H. Hasenauer](https://doi.org/10.1002/joc.4436) and further developed by W. Rammer, C. Pucher & M. Neumann from the University of Natural Resources and Life Sciences, Vienna, Austria. Please, check [this document](https://github.com/VeruGHub/easyclimate/tree/master/inst/Description_Evaluation_Validation_Downscaled_Climate_Data_v2.pdf) for more details on the development and characteristics of the climatic dataset.


## Installation

``` r
# install.packages("remotes")
remotes::install_github("VeruGHub/easyclimate")
```


## Examples

To obtain a data frame of daily climatic values for point coordinates:

```{r example}
library(easyclimate)

coords <- matrix(c(-5.36, 37.40), ncol = 2)

prec <- get_daily_climate(coords, 
                          period = "2001-01-01:2001-01-03", 
                          climatic_var = "Prcp")
```

```{r echo=FALSE}
kable(prec)
```

<br>

To obtain a (multi-layer) raster of daily climatic values for an area:

```{r message=FALSE}
library(terra)

coords_poly <- vect("POLYGON ((-4.5 41, -4.5 40.5, -5 40.5, -5 41))")

ras_tmax <- get_daily_climate(
  coords_poly,
  period = c("2012-01-01", "2012-08-01"),
  climatic_var = "Tmax",
  output = "raster" # return raster
  )

ras_tmax <- ras_tmax/100

par(mfrow = c(1, 2))
terra::plot(ras_tmax, 1, col = rev(heat.colors(20)), type = "continuous", smooth = TRUE, range = c(10, 35), legend = FALSE, mar=c(4, 2, 4, 2), main = "January 1 2012")
terra::plot(ras_tmax, 2, col = rev(heat.colors(20)), type = "continuous", smooth = TRUE, range = c(10, 35), mar = c(4, 1, 4, 3), main = "August 1 2012")

```

<br>
Visit the articles of the package website for more extended tutorials!

<br>

## CITATION

If you use easyclimate, please cite both the data source and the package as:

```{r echo=FALSE, results='asis'}
print(citation("easyclimate"), style = "text")
```



