---
title: "easyclimate: Analysing the climate of a given period using data frames"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{easyclimate: Analysing the climate of a given period using data frames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, collapse = TRUE, warning = FALSE, message = FALSE)
```

The objective of `easyclimate` is to easily download daily climate data for a given set of points or polygons within Europe.

# Installation

You can download and install the development version of `easyclimate` from github using:

```{r, eval = FALSE}
remotes::install_github("VeruGHub/easyclimate")
```

# An example using data frames as input and output

As the name suggests `easyclimate` is very easy to use. Simply create a data frame containing at least two columns named `lon` and `lat` (longitude and latitude coordinates, respectively), and download the daily climate data for that location for the specified period (e.g. 2008-2010).

```{r, results = "hide"}
library(easyclimate)
library(tidyverse)
library(viridis)
library(patchwork)
library(here)
```

```{r}
coords <- data.frame(
  lon = -2.41135, lat = 43.0329
  )

Sys.time()

ex_prcp <- get_daily_climate(
  coords, 
  period = 2008 : 2010, 
  climatic_var = "Prcp",
  output = "df"
)

ex_tmin <- get_daily_climate(
  coords,
  period = 2008 : 2010,
  climatic_var = "Tmin",
  output = "df"
)

ex_tmax <- get_daily_climate(
  coords,
  period = 2008 : 2010,
  climatic_var = "Tmax",
  output = "df"
)

Sys.time()
```

Note that temperature and precipitation values are in ºC\*100 and mm\*100, respectively, to avoid floating values. You can mutate them and join all the data frames.

```{r}
glimpse(ex_prcp)
ex_prcp_date <- ex_prcp %>%
  mutate(
    prcp = Prcp / 100,
    date = as.Date(date),
    month = months(date),
    year = format(date, format = "%y")
         )

ex_min_date <- ex_tmin %>%
  mutate(
    tmin = Tmin / 100,
    date = as.Date(date)
  ) %>%
  select(
    tmin, date
  )

ex_max_date <- ex_tmax %>%
  mutate(
    tmax = Tmax / 100,
    date = as.Date(date)
  ) %>%
  select(
    tmax, date
  )

clim_join <- full_join(
  ex_prcp_date, ex_min_date, by = "date"
  ) %>%
  full_join(
    ex_max_date, by = "date"
    ) %>%
  select(
    !c(ID_coords, x, y, Prcp)
  ) %>%
  relocate(
    lon, lat, date, year, month, prcp, tmin, tmax
    )
```

Finally, you can calculate different climate variables (e.g. precipitation and maximum temperature per month, and late-spring frost days) and visualise the results.

```{r, fig.width = 8, fig.height = 6}
gg_clim <- function(
  df,
  x,
  y,
  grouping,
  virop,
  title
  ){
  ggplot(df, aes(x = .data[[x]], y = .data[[y]], color = .data[[y]])) +
  geom_point() +
  facet_wrap(~year) +
  geom_line(aes(x = .data[[x]], y = .data[[y]], group = year)) +
  scale_colour_viridis(option = virop, direction = -1) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    axis.title = element_blank(),
    panel.grid.major = element_line(colour = "grey90", size = 0.5),
    panel.background = element_blank(),
    plot.title = element_text(size = 10)
  )  +
  ggtitle(title)
}

months <- c(
  "February", "March", "April",
  "May", "June", "July", "August",
  "September", "October", "November", "December"
  )

# precipitation per month
gg_prcp <- clim_join %>%
  mutate(
    month = fct_relevel(month, months, after = 5)) %>%
  group_by(month, year) %>%
  mutate(prcp = sum(prcp)) %>% 
  gg_clim(
    x = "month",
    y = "prcp",
    virop = "mako",
    title = "Precipitation per month"
  )

# maximum temperature per month
gg_tmax <- clim_join %>%
  mutate(
    month = fct_relevel(month, months, after = 5)) %>%
  group_by(month, year) %>%
  filter(tmax == max(tmax)) %>%
  gg_clim(
    x = "month",
    y = "tmax",
    virop = "rocket",
    title = "Maximum temperature per month"
  )

months <- c( "March", "April",
             "May", "June")

# late-spring frost days
gg_tmin <- clim_join %>%
  filter(
    tmin < 0 & (
      month == "March" |
      month == "April" |
      month == "May" |
      month == "June")
  ) %>%
  mutate(n.days = sum(n())) %>%
  gg_clim(
    x = "month",
    y = "n.days",
    virop = "mako",
    #br = c(1, 5, 10),
    title = "Late-spring frost days"
  ) +
  scale_y_continuous(breaks = c(1, 5, 10))

# visualise and save all plots together
gg_prcp / gg_tmax / (plot_spacer() + plot_spacer() + gg_tmin) +
  plot_annotation(
  title = "Oñati (Gipuzkoa, Spain) climate 2008-2010",
)

ggsave(
  here("clim-08-10.png"), 
  width = 8, height = 6
)
```

# Learn more

`easyclimate` can handle different types of input and output objects or download data from multiple locations and periods. Check out the vignettes with more information on the other options in this package here:
