---
title: "Analysing the climate at spatial points for a given period"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Analysing the climate at spatial points for a given period}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 6, fig.height = 4, cache = TRUE)
```

With `easyclimate` you can easily download daily climate data for a given set of points or polygons within Europe. To download and install the latest version of `easyclimate` from github follow the instructions in  https://github.com/VeruGHub/easyclimate 

<br>

In this tutorial we will work through the basics of using easyclimate with coordinate points. You can enter coordinates as a `data.frame`, `matrix`, `sf` or `SpatVector` object. At the end we will have a data frame with climate variables for each point.

<br>

# Example 1: Introducing coordinates as a data frame

First, specify longitude and latitude coordinates in a data frame with the column names `lon` and `lat`. Here we are simulating coordinates for three random sites in southern Spain

```{r}
library(easyclimate)
library(ggplot2)
library(dplyr)

coords <- data.frame(
  lon = rnorm(3, mean = -5.36, sd = 0.3), 
  lat = rnorm(3, mean = 37.40, sd = 0.3)
  )

ggplot() +
  borders(regions = c("Spain", "Portugal", "France")) +
	geom_point(data = coords, aes(x = lon, y = lat)) +
  coord_fixed(xlim = c(-10, 2), ylim = c(36, 44), ratio = 1.3) + 
	xlab("Longitude") +
	ylab("Latitude") +
  theme_bw()

```

<br>

Now, download the climatic data for the selected locations. All you have to do is use the function `get_daily_climate`, specifying the period (e.g. `2008-05-25` for a single day or `2008:2010` for several years), and the variables to be downloaded (precipitation `Prcp`, minimum temperature `Tmin` or maximum temperature `Tmax`).

```{r}

Sys.time() # to know how much time it takes to download

daily <- get_daily_climate(
  coords = coords, 
  period = 2008:2010, 
  climatic_var = c("Prcp","Tmin","Tmax"))

Sys.time()

kable(head(daily))

```

Here we extract different components of the date. 

```{r}

daily <- daily |> 
  mutate(
    date = as.Date(date),
    month = months(date),
    year = format(date, format = "%y")
    ) 

```

<br>

Finally, you can visualize the daily climate results. For example, let's plot the precipitation for one of the sites:

```{r}
clim_site1 <- daily |> 
  filter(ID_coords == 1)

ggplot(clim_site1) +
  geom_line(aes(x = date, y = Prcp), colour = "steelblue") +
  labs(x = "Date", y = "Daily precipitation (mm)") +
  theme_bw()
```

<br>

Or calculate the daily mean temperature and plot it against `tmin` and `tmax`:

```{r}
library(tidyr)

temp_long <- daily |> 
  mutate(Tmean = (Tmin + Tmax) / 2) |> 
  pivot_longer(
    cols = c("Tmin", "Tmax", "Tmean"),
    names_to = "temp_vars",
    values_to = "temp_values")

ggplot(temp_long, aes(x = factor(ID_coords), y = temp_values, fill = temp_vars, color = temp_vars)) +
  geom_violin(size = 1, alpha = .7) +
    scale_fill_manual(values = c("#C54A52", "#FAFBC5", "#4B8AB8")) +
  scale_color_manual(values = c("#C54A52", "#FAFBC5", "#4B8AB8")) +
  ylab("Temperature (ºC)") + xlab("") +
  theme_bw()

ggplot(temp_long, aes(x = date, y = temp_values, color = temp_vars)) + 
  geom_point(alpha = .3) +
  scale_color_manual(name = "Variables",
                     values = c("#C54A52", "#FAFBC5", "#4B8AB8"),
                     guide = guide_legend(override.aes = list(alpha = 1))) +
  ylab("Temperature (ºC)") + xlab("Date") +
  theme_bw()

```


<br>
<br>

# Example 2: Introducing coordinates as a matrix

`easyclimate` handles different input data, try now with matrices! 

Here we are retrieving daily precipitation data for a single year (2008).

```{r}

coords_mat <- as.matrix(coords)

Sys.time()

mat_prcp <- get_daily_climate( 
  coords = coords_mat, 
  period = 2008, # single year
  climatic_var = "Prcp"
)

Sys.time()

kable(head(mat_prcp))

mat_prcp <- mat_prcp |> 
  mutate(
    date = as.Date(date),
    month = months(date),
    year = format(date, format = "%y")
  ) |> 
  relocate(lon, lat, date, year, month, Prcp)

ggplot(mat_prcp, aes(x = date, y = Prcp, color = Prcp)) +
  geom_point() +
  scale_color_continuous(name = "Precipitation (mm)") +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim = c(0, 60)) +
  ylab("Daily precipitation (mm)") + xlab("Date") +
  theme_bw()

month_name <- format(ISOdate(2021, 1:12, 1), "%B")

mat_prcp |> 
  mutate(month = factor(month, rev(month_name))) |> 
  ggplot(aes(x = month, y = Prcp, color = month)) +
  geom_jitter(size = 2, alpha = .3, width = .3, show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0)) +
  coord_flip(ylim = c(0, 60)) +
  ylab("Daily precipitation (mm)") + xlab("Month") +
  theme_bw()

```


<br>
<br>

# Example 3: Introducing coordinates as simple feature objects

Here we introduce coordinates as a `sf` object, and retrieve minimum temperature for a single day (1 January 2001).


```{r}
library(sf)

coords_sf <- st_as_sf(
  coords, 
  coords = c("lon", "lat")
  )

sf_tmin <- get_daily_climate(
  coords = coords_sf, 
  period = "2001-01-01", # single day
  climatic_var = "Tmin"
  ) 

ggplot() +
  borders(regions = c("Spain", "Portugal", "France")) +
	geom_point(data = sf_tmin, aes(x = lon, y = lat, color = Tmin), size = 2) + 
  coord_fixed(xlim = c(-10, 2), ylim = c(36, 44), ratio = 1.3) +  
  scale_color_gradient2(name = "Minimum\ntemperature (ºC)",
                        low = "#4B8AB8", mid = "#FAFBC5", high = "#C54A52",
                        midpoint = mean(sf_tmin$Tmin)) +
	ylab("Latitude") + xlab("Longitude") +
  theme_bw()

```

<br>

## Learn more

Now you know how to obtain a data frame with different climatic variables with `easyclimate`, using point coordinates in different formats and downloading data for multiple locations and periods. Check out this [other vignette](https://verughub.github.io/easyclimate/articles/polygons-raster.html) if you need to extract the data of a complete area.
