---
title: "Analysing the climate at spatial points for a given period"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Analysing the climate at spatial points for a given period}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 6, fig.height = 4, cache = TRUE)
```

With `easyclimate` you can easily download daily climate data for a given set of points or polygons within Europe. To download and install the latest version of `easyclimate` from github follow the instructions in  https://github.com/VeruGHub/easyclimate 

<br>

In this tutorial we will work through the basics of using easyclimate with coordinate points. You can enter coordinates as a `data.frame`, `matrix`, or `sf` object. At the end we will have a data frame with climate variables for each point.

<br>

# Example 1: Introducing coordinates as a data frame

First, specify longitude and latitude coordinates in a data frame with the column names `lon` and `lat`. Here we are simulating coordinates for three random sites in southern Spain

```{r}
library(easyclimate)
library(ggplot2)
library(dplyr)

coords <- data.frame(
  lon = rnorm(3, mean = -5.36, sd = 0.3), 
  lat = rnorm(3, mean = 37.40, sd = 0.3)
  )

ggplot() +
  borders(regions = c("Spain", "Portugal", "France")) +
	geom_point(data = coords, aes(x = lon, y = lat)) +
  coord_fixed(xlim = c(-10, 2), ylim = c(36, 44), ratio = 1.3) + 
	xlab("Longitude") +
	ylab("Latitude") +
  theme_bw()

```

<br>

Now, download the climatic data for the selected locations. It is very simple, all you have to do is use the function `get_daily_climate`, specify the period (e.g. `2008-05-25` for a single day or `2008:2010` for several years), and the variables to be downloaded (precipitation `Prcp`, minimum temperature `Tmin` or maximum temperature `Tmax`).

```{r}

Sys.time() # to know how much it takes to download

df_prcp <- get_daily_climate(
  coords = coords, 
  period = 2008:2010, 
  climatic_var = "Prcp"
)

df_tmin <- get_daily_climate(
  coords = coords,
  period = 2008:2010,
  climatic_var = "Tmin"
)

df_tmax <- get_daily_climate(
  coords = coords,
  period = 2008:2010,
  climatic_var = "Tmax"
)

Sys.time()

head(df_prcp[, c("lon", "lat", "date", "Prcp")])

```

Temperature and precipitation units are ºC\*100 and mm\*100, respectively, to avoid floating values. You can change the units using the mutate function and join all the variables in a data frame. 

```{r}

prcp_date <- df_prcp %>%
  mutate(prcp = Prcp / 100) %>% 
  select(ID_coords, lon, lat, date, prcp)
  
tmin_date <- df_tmin %>%
  mutate(tmin = Tmin / 100) %>% 
  select(ID_coords, date, tmin)

tmax_date <- df_tmax %>%
  mutate(tmax = Tmax / 100) %>% 
  select(ID_coords, date, tmax)


clim_join <- full_join(prcp_date, tmin_date, by = c("ID_coords", "date")) %>%
  full_join(tmax_date, by =  c("ID_coords", "date")) %>%
  mutate(
    date = as.Date(date),
    month = months(date),
    year = format(date, format = "%y")
  ) %>%
  select(ID_coords, lon, lat, date, year, month, prcp, tmin, tmax)


kable(head(clim_join))

```

<br>

Finally, you can visualize the daily climate results. For example, let's plot the precipitation for one of the sites:

```{r}
clim_site1 <- clim_join %>% 
  filter(ID_coords == 1)

ggplot(clim_site1) +
  geom_line(aes(x = date, y = prcp), colour = "steelblue") +
  labs(x = "Date", y = "Daily precipitation (mm)") +
  theme_bw()
```

<br>

Or calculate the daily mean temperature and plot it against `tmin` and `tmax`:

```{r}
library(tidyr)

clim_df <- clim_join %>%
  mutate(tmean = (tmin + tmax) / 2) %>%
  pivot_longer(
    cols = c("tmin", "tmax", "tmean"),
    names_to = "temp_vars",
    values_to = "temp_values")

ggplot(clim_df, aes(x = temp_vars, y = temp_values, color = temp_vars)) +
  geom_violin(fill = "gray80", size = 1, alpha = .5, show.legend = FALSE) +
  coord_flip() +
  ylab("Temperature (ºC)") + xlab("") +
  theme_bw()

ggplot(clim_df, aes(x = date, y = temp_values, color = temp_vars)) + 
  geom_point(alpha = .1) +
  scale_color_discrete(name = "Variables", 
                       guide = guide_legend(override.aes = list(alpha = 1))) +
  ylab("Temperature (ºC)") + xlab("Date") +
  theme_bw()

```


<br>
<br>

# Example 2: Introducing coordinates as a matrix

`easyclimate` handles different input data, try now with matrices! 

Here we are retrieving daily precipitation data for a single year (2008).

```{r}

coords_mat <- as.matrix(coords)

Sys.time()

mat_prcp <- get_daily_climate( 
  coords = coords_mat, 
  period = 2008, # single year
  climatic_var = "Prcp"
)

Sys.time()

head(mat_prcp)

clim_mat <- mat_prcp %>%
  mutate(
    prcp = Prcp / 100,
    date = as.Date(date),
    month = months(date),
    year = format(date, format = "%y")
  ) %>%
  relocate(x, y, date, year, month, prcp)

ggplot(clim_mat, aes(x = date, y = prcp, color = prcp)) +
  geom_point() +
  scale_color_continuous(name = "Precipitation (mm)") +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim = c(0, 60)) +
  ylab("Daily precipitation (mm)") + xlab("Date") +
  theme_bw()

month_name <- format(ISOdate(2021, 1:12, 1), "%B")

clim_mat %>%
  mutate(month = factor(month, rev(month_name))) %>%
  ggplot(aes(x = month, y = prcp, color = month)) +
  geom_jitter(size = 2, alpha = .3, width = .3, show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0)) +
  coord_flip(ylim = c(0, 60)) +
  ylab("Daily precipitation (mm)") + xlab("Month") +
  theme_bw()

```


<br>
<br>

# Example 3: Introducing coordinates as simple feature objects

Here we introduce coordinates as a `sf` object, and retrieve minimum temperature for a single day (1 January 2001).


```{r}
library(sf)

coords_sf <- st_as_sf(
  coords, 
  coords = c("lon", "lat")
  )

sf_tmin <- get_daily_climate(
  coords = coords_sf, 
  period = "2001-01-01", # single day
  climatic_var = "Tmin"
  ) 

clim_sf <- sf_tmin %>%
  mutate(tmin = Tmin / 100)

ggplot() +
  borders(regions = c("Spain", "Portugal")) +
	geom_point(data = clim_sf, aes(x = x, y = y, color = tmin), size = 2) + 
  coord_fixed(xlim = c(-10, 2), ylim = c(36, 44), ratio = 1.3) +  
  scale_color_continuous(type = "viridis", name = "Minimum\ntemperature (ºC)") +
	ylab("Latitude") + xlab("Longitude") +
  theme_bw()

```

<br>

## Learn more

Now you know how to extract a data frame with different climatic variables with `easyclimate`, using point coordinates in different formats and downloading data for multiple locations and periods. Check out this [other vignette](https://verughub.github.io/easyclimate/articles/polygons-raster.html) if you need to extract the data of a complete area.
