---
title: "Analysing the climate of an area for a given period"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Analysing the climate of an area for a given period}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 6, fig.height = 4, cache = TRUE)
```

With `easyclimate` you can easily download daily climate data for a given set of points or polygons within Europe. To download and install the latest version of `easyclimate` from github follow the instructions in https://github.com/VeruGHub/easyclimate 

<br>

In this tutorial we will work through the basics of using `easyclimate` with a spatial polygon.

If you wish to download the climatic data of a specific region, you need to specify at least four corners of the polygon including the area and specify the type of output you want to obtain (i.e. a data frame - `df` or a raster - `raster`). You can also provide the polygons of interest in a `sf` object.

```{r}
library(easyclimate)
library(terra)

coords_t <- vect("POLYGON ((-4.5 41, -4.5 40.5, -5 40.5, -5 41))")

Sys.time() # to know how much it takes to download

df_tmax <- get_daily_climate(
  coords_t,
  period = c("2012-01-01", "2012-08-01"),
  climatic_var = "Tmax",
  output = "df" # return dataframe
  )

Sys.time()

head(df_tmax)

```

```{r}
library(dplyr)

clim_df <- df_tmax %>%
  mutate(
    date = as.Date(date)
  ) 

```

Then, you can visualize the results and compare both dates

```{r}

library(ggplot2)

tapply(clim_df$Tmax, clim_df$date, summary)

ggplot() +
	geom_raster(data = clim_df,
		aes(x = lon, y = lat, fill = Tmax)) +
  scale_fill_continuous(type = "viridis", name = "Maximum\ntemperature (ºC)") +
  facet_wrap(~date) +
	ylab("Latitude") + xlab("Longitude") +
  theme_bw()

```

You can get a (multi-layer) raster directly as output, if you specify `output = raster`:

```{r}

Sys.time()

ras_tmax <- get_daily_climate(
  coords_t,
  period = c("2012-01-01", "2012-08-01"),
  climatic_var = "Tmax",
  output = "raster" # return raster
  )

Sys.time()

ras_tmax

par(mfrow = c(1, 2))
terra::plot(ras_tmax, 1, col = rev(heat.colors(20)), type = "continuous", smooth = TRUE, range = c(10, 35), legend = FALSE, mar=c(4, 2, 4, 2), main = "January 1 2012")
terra::plot(ras_tmax, 2, col = rev(heat.colors(20)), type = "continuous", smooth = TRUE, range = c(10, 35), mar = c(4, 1, 4, 3), main = "August 1 2012") 

```

<br>

## Learn more

Now you know how to extract climatic variables with `easyclimate`, for a specific area. Check out this [other vignette](https://verughub.github.io/easyclimate/articles/points-df-mat-sf.html) if you need to extract the data for specific points.
