---
title: "Analysing the climate of an area for a given period"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Analysing the climate of an area for a given period}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 6, fig.height = 4, cache = TRUE)
```

`easyclimate` aims to easily download daily climate data for a given set of points or polygons within Europe. You can download and install the latest version of `easyclimate` from github: https://github.com/VeruGHub/easyclimate 

<br>

In this tutorial we will work through the basics of using `easyclimate` with a spatial polygon.

If you wish to download the climatic data of a specific region, you need to specify at least four corners of the polygon including the area and specify the type of output you want to obtain (i.e. a dataframe - "df" or a raster - "raster"). You can also provide the polygons of interest in a `sf` object.

```{r}
library(easyclimate)
library(terra)

coords_t <- vect("POLYGON ((-4.5 41, -4.5 40.5, -5 40.5, -5 41))")

Sys.time() # to know how much it takes to download

df_tmax <- get_daily_climate(
  coords_t,
  period = c("2012-01-01", "2012-08-01"),
  climatic_var = "Tmax",
  output = "df" # return dataframe
  )

Sys.time()

head(df_tmax)

```

Note that temperature is given as ºC\*100 to avoid floating values. You can mutate this variable as follows:

```{r}
library(dplyr)

clim_df <- df_tmax %>%
  mutate(
    tmax = Tmax / 100,
    date = as.Date(date)
  ) 

```

Then, you can visualize the results and compare both dates

```{r}

library(ggplot2)

tapply(clim_df$tmax, clim_df$date, summary)

ggplot() +
	geom_raster(data = clim_df,
		aes(x = x, y = y, fill = tmax)) +
  scale_fill_continuous(type = "viridis", name = "Maximum\ntemperature (ºC)") +
  facet_wrap(~date) +
	ylab("Latitude") + xlab("Longitude") +
  theme_bw()

```

You can directly get a (multi-layer) raster as output, if you specify output = "raster":

```{r}

Sys.time()

ras_tmax <- get_daily_climate(
  coords_t,
  period = c("2012-01-01", "2012-08-01"),
  climatic_var = "Tmax",
  output = "raster" # return raster
  )

Sys.time()

ras_tmax

ras_tmax <- ras_tmax/100
  
plot(ras_tmax, col = rev(heat.colors(7))) 

```

<br>

## Learn more

Now you know how to extract climatic variables with `easyclimate`, for a specific area. Check out this [other vignette](https://verughub.github.io/easyclimate/articles/points-df-mat-sf.html) if you need to extract the data for specific points.
